A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-10-22 JNABRev20231018/JNAB2review-response.txt

I appreciate the effort that the authors have made in this revision. I believe the paper has
been considerably improved.

>> Thanks. Appreciated.

I think it would useful to expand a bit on some of the comments in the section “An illustrative
example”. In particular, the choice of starting estimates, at least for the third formulation of
the model, doesn’t make sense to me.

>> Added explanatory comment.
>> <!-- 231018: attempt to address comments in 2report1 on modeling vs least squares -->
>> We thank one of the referees for expressing the viewpoint that a nonlinear regression
>> approach would use background knowledge and graphs to obtain an understanding of
>> the general magnitude of the parameters. Indeed, the "visual fitting" approach
>> (@nash1996nonlinear) mentioned later accords with this viewpoint, as does use
>> of the **Logistic3T** variant of the model, as well as discussions in @Ross90, 
>> @Seber89Wild, @bateswatts, and @Gallant87.
>> 
>> Our emphasis on the software resilience to starting parameters that are, in the
>> sense of statistical modeling, "silly" comes from over half a century of dealing
>> with users whose interests and understanding are very far from those of statistical
>> modelers. Thus we seek methods and codes that obtain reasonable answers under
>> highly unfavourable conditions. Nevertheless, a proper approach to 
>> nonlinear modeling is to apply all available knowledge to the task.
>> <!-- end 231018 work -->

@@

>> Agreed, and we believe we have now done so. To be honest, the original c(1,1,1) parameters
>> of the 1970s were likely pure laziness and quick to type.

On page 4 the authors state that if a selfStart model is not available nls() sets the starting
values to 1. That may be the case now but the original intention was certainly that a user
calling nls() would provide starting estimates suitable to the model and data.

Also on page 4 I don’t understand the comment in the hobbs.res function that This variant
uses looping. I don’t see any explicit loops. There will naturally be looping in the C code
that implements the arithmetic operations on vectors but that is inevitable.

>> The comment was a left-over from an earlier variant of the functions and should have been
>> removed. Done. Thanks for spotting this.
>> We also noted that the model is UNscaled, and edited the comment.
>> An explanatory comment has been added.
>> 
>> The particular form of these explicit residual and Jacobian functions comes from
>> their translation from BASIC codes of the 1970s, as adapted in @jnmws87. The use
>> of the Laplace form of the residual and the inclusion of data within the functions
>> reflects choices of that era that are at odds with current practice. Some users
>> do still want or need to provide problems as explicit functions, particularly
>> for problems that are NOT regressions. For example, the Rosenbrock banana-valley
>> test problem can be provided this way, where the two "residuals" are different
>> functional forms.

### 2023-10-22 JNABRev20231018/ComparingNLStools.Rmd
various updates for JNAB revision

draft: true
author:
    - name: John C. Nash
      affiliation: retired professor, University of Ottawa
      address:
      - Telfer School of Management
      - Ottawa ON Canada K1N 6N5
      orcid: 0000-0002-2762-8039
      email: profjcnash@gmail.com
    - name: Arkajyoti Bhattacharjee
      affiliation: Indian Institute of Technology
      address:
      - Department of Mathematics and Statistics
      - Kanpur
      email: arkastat98@gmail.com
type: package
output: 
  rjtools::rjournal_web_article:
      self_contained: yes
      toc: no
bibliography: ./ComparingNLStools.bib
---

@@

`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly **not** about modeling, and
the input and output are quite different from class `nls` methods. However, there do
not appear to be other R packages with these capabilities.

@@

<!-- 231018: attempt to address comments in 2report1 on modeling vs least squares -->
We thank one of the referees for expressing the viewpoint that a nonlinear regression
approach would use background knowledge and graphs to obtain an understanding of
the general magnitude of the parameters. Indeed, the "visual fitting" approach
(@nash1996nonlinear) mentioned later accords with this viewpoint, as does use
of the **Logistic3T** variant of the model.

>>>>

draft: true
author:
    - name: John C. Nash
      affiliation: retired professor, University of Ottawa
      address:
      - Telfer School of Management
      - Ottawa ON Canada K1N 6N5
      orcid: 0000-0002-2762-8039
      email: profjcnash@gmail.com
    - name: Arkajyoti Bhattacharjee
      affiliation: Indian Institute of Technology
      address:
      - Department of Mathematics and Statistics
      - Kanpur
      email: arkastat98@gmail.com
type: package
output: 
  rjtools::rjournal_web_article:
      self_contained: yes
      toc: no
  # pdf_document
bibliography: ./ComparingNLStools.bib
---

@@

`nlsic` (@nlsic22) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly **not** about modeling, and
the input and output are quite different from class `nls` methods. However, there do
not appear to be other R packages with these capabilities.

@@

<!-- 231018: attempt to address comments in 2report1 on modeling vs least squares -->
We thank one of the referees for expressing the viewpoint that a nonlinear regression
approach would use background knowledge and graphs to obtain an understanding of
the general magnitude of the parameters. Indeed, the "visual fitting" approach
(@nash1996nonlinear) mentioned later accords with this viewpoint, as does use
of the **Logistic3T** variant of the model, as well as discussions in @Ross90, 
@Seber89Wild, @bateswatts, and @Gallant87.

### 2023-10-22 JNABRev20231018/ComparingNLStools.tex
various updates for JNAB revision

\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
Torvisco, Fernandez, and Sanchez (2018) for fitting separable nonlinear regression curves. The
special class of problem for which this package is intended is an important and
difficult one. No vignette is provided, unfortunately.

@@

We thank one of the referees for expressing the viewpoint that a nonlinear regression
approach would use background knowledge and graphs to obtain an understanding of
the general magnitude of the parameters. Indeed, the ``visual fitting'' approach
(John C. Nash and Velleman (1996)) mentioned later accords with this viewpoint, as does use
of the \textbf{Logistic3T} variant of the model.

@@

@@

@@

\bibitem[\citeproctext]{ref-NISTnls-manual}
National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.

@@

\bibitem[\citeproctext]{ref-nlsic-manual}
Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.

\bibitem[\citeproctext]{ref-Torvisco2018}
Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias Fernandez, and Javier Cabello Sanchez. 2018. {``A New Algorithm to Fit Exponential Decays Without Initial Guess.''} {Faculty of Sciences and Mathematics, University of Niˇs, Serbia}. \url{https://doi.org/doi.org/10.2298/FIL1812233T}.

\bibitem[\citeproctext]{ref-R-crsnls-manual}
Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}. \url{https://CRAN.R-project.org/package=crsnls}.

>>>>

\texttt{NISTnls} ((ed.) (2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
Torvisco, Rodrıguez-Arias, and Sanchez (2018) for fitting separable nonlinear regression curves. The
special class of problem for which this package is intended is an important and
difficult one. No vignette is provided, unfortunately.

@@

We thank one of the referees for expressing the viewpoint that a nonlinear regression
approach would use background knowledge and graphs to obtain an understanding of
the general magnitude of the parameters. Indeed, the ``visual fitting'' approach
(John C. Nash and Velleman (1996)) mentioned later accords with this viewpoint, as does use
of the \textbf{Logistic3T} variant of the model, as well as discussions in Ross (1990),
Seber and Wild (1989), D. M. Bates and Watts (1988), and Gallant (1987).

@@

\bibitem[\citeproctext]{ref-NISTnls-manual}
(ed.), Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.

@@

\bibitem[\citeproctext]{ref-Gallant87}
Gallant, A. Ronald. 1987. \emph{Nonlinear Statistical Models}. Wiley.

@@

@@

\bibitem[\citeproctext]{ref-Ross90}
Ross, Gavin J. S. 1990. \emph{Nonlinear Estimation.} New York etc.: Springer-Verlag.

\bibitem[\citeproctext]{ref-Seber89Wild}
Seber, G. A. F., and C. J. Wild. 1989. \emph{{Nonlinear regression}}. New York: Wiley.

\bibitem[\citeproctext]{ref-nlsic22}
Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.

\bibitem[\citeproctext]{ref-Torvisco2018}
Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias, and Javier Cabello Sanchez. 2018. {``A New Algorithm to Fit Exponential Decays Without Initial Guess.''} {Faculty of Sciences and Mathematics, University of Niˇs, Serbia}. \url{https://doi.org/doi.org/10.2298/FIL1812233T}.

\bibitem[\citeproctext]{ref-R-crsnls-manual}
Tvrdík, Josef. 2016. \emph{{crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}}. \url{https://CRAN.R-project.org/package=crsnls}.
